%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Numerical Linear Algebra class 2022 
% Sheet 6
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{Sheet}[to be handed in until December 2, 2022, 2pm.]
  \label{sheet6}

  \begin{Problem}
    Prove that a matrix iteration converges \textit{if and only if}
    the spectral radius of the iteration matrix is less than 1.
    \textit{Hint:} You can use Lemma 2.2.10 in the lecture notes.
  \end{Problem}

  \begin{Problem}
    \label{sheet6:problem2}
    Consider the $n\times n$ tridiagonal matrix
    \begin{gather*}
      \matt_{\alpha} =
      \begin{pmatrix}
        \alpha &     -1 &        &        & \\
            -1 & \alpha &     -1 &        & \\
               & \ddots & \ddots & \ddots & \\
               &        &     -1 & \alpha &     -1 \\
               &        &        &     -1 & \alpha
      \end{pmatrix}
    \end{gather*}
    where $\alpha$ is a real parameter. Verify that the eigenvalues of
    $\matt_{\alpha}$ are given by
    \begin{gather*}
      \lambda_j = \alpha - 2\cos(j\theta),
    \end{gather*}
    where
    \begin{gather*}
      \theta = \frac\pi{n+1},
    \end{gather*}
    and the eigenvectors associated with each $\lambda_j$ are given by
    \begin{gather*}
      \vv_j = \left(
        \sin(j\theta),\ \sin(2j\theta),\ \ldots\ ,\ \sin(nj\theta)
      \right)^T.
    \end{gather*}
    What are the conditions on $\alpha$, such that the matrix
    $\matt_{\alpha}$ becomes positive-definite?
  \end{Problem}

  \begin{Problem}
    \label{sheet6:problem3}
    Let $\mata$ be a real symmetric positive-definite $n\times n$
    matrix and $\vb$ a vector in $\R^n$. Consider the Richardson
    iteration defined as
    \begin{gather*}
      \vx_{k+1}
      = \left(\id-\omega\mata \right)\vx_k + \omega\vb
      = \vx_k - \omega\left(\mata\vx_k - \vb \right)
    \end{gather*}
    where $\omega$ is a constant real parameter.
    \begin{enumerate}[(a)]
    \item Show that the sequence of iterates $\vx_k$ generated by the
      Richardson iteration converges to $\mata^{-1}\vb$ for any
      initial vector $\vx_0$ \textit{if and only if}
      \begin{gather*}
        0 < \omega < \frac{2}{\lambda_{\max}}.
      \end{gather*}
    \item Show that the optimal value of $\omega$ is
      \begin{gather*}
        \omega_{opt} = \frac2{\lambda_{\max} + \lambda_{\min}}.
      \end{gather*}
      Moreover
      \begin{gather*}
        \norm{\id-\omega_{opt}\mata}_2
        = \frac{\lambda_{\max} - \lambda_{\min}}{\lambda_{\max} + \lambda_{\min}}
        = \frac{\cond_2(\mata)-1}{\cond_2(\mata)+1},
      \end{gather*}
      where $\cond_2(\mata) = \frac{\lambda_{\max}}{\lambda_{\min}}$.
    \end{enumerate}
  \end{Problem}

  \begin{Problem}
    Apply the obtained results of Problem \ref{sheet6:problem3} to a
    one-dimensional Laplace problem. To this end, take matrix
    $\matt_{\alpha}$ from Problem \ref{sheet6:problem2} with
    $\alpha = 2$ and work on the following tasks.
    \begin{enumerate}[(a)]
    \item Will the Richardson iteration coverge for this matrix? If
      so, what will be its covergence rate?
      \begin{todo}
        How do you have to choose $\omega$, such that the Richardson
        iteration converges? What is the best choice?
      \end{todo}
    \item How many iterations are needed to reduce the error by a
      factor of $10^{-6}$ depending on the number $n$ of dicretization
      points?
    \item Estimate the cost of a single iteration (depending on $n$,
      use $\mathcal{O}(\cdot)$ notation).
    \item Estimate the cost of reducing the error by a given factor
      (depending on $n$, use $\mathcal{O}(\cdot)$ notation).
    \end{enumerate}
  \end{Problem}

\end{Sheet}


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End: 
